Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-24 paper/geometry-twocolumn.tex
small changes

The $n$-tuple of numbers is the basic data structure for programs that
manipulate geometric data. Pairs, i.e., $2$-tuples, 
%The basic data structure used in programs that manipulate geometric data is
%the $n$-tuple of numbers. In the 2-dimensional case, tuples 
$\vec{v} = (x,y)$ serve double duty, representing
%are called upon to represent both 
both \emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right.  Despite their common
representation, points and vectors are very different, and
distinguishing between them is the key feature of \emph{affine
  geometry} (see, for example, Chapter 2 of Gallier's book
\cite{gallier11geometric}). Nevertheless, computational geometry
libraries traditionally either leave it to the programmer to maintain
the distinction between points and vectors, or else use different
abstract types for points and vectors to enforce it.
%(one might say that the standard mathematical formulation of affine
%spaces uses the abstract types approach). 
In this paper we investigate a more sophisticated approach based on 
types indexed by change of origin transformations.

@@

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
we assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The (simplified) index-erasure and relation interpretations
of $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{\rho'} \sepbar \rho' \in
    \extends{\rho}{t\mathord:\SynTransl{2}} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
depends on which (possibly non-compositional) interpretation of
relational environments we use. For now we can think of environments
as mappings from free variables of sort $\SynTransl{2}$ to the group
$\Transl{2}$, and of extension as adding extra components to
mappings. The index-erasure and relational interpretations are given
formally in Sections~\ref{sec:index-erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\ref{sec:relational-semantics}. %, respectively.

>>>>

The basic data structure used in programs that manipulate geometric data is
the $n$-tuple of numbers. In the 2-dimensional case, tuples 
$\vec{v} = (x,y)$ serve double duty, representing
%are called upon to represent both 
both \emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right.  Despite their common
representation, points and vectors are very different, and
distinguishing between them is the key feature of \emph{affine
  geometry} (see, for example, Chapter 2 of Gallier's book
\cite{gallier11geometric}). Nevertheless, computational geometry
libraries traditionally either leave it to the programmer to maintain
the distinction between points and vectors, or else use different
abstract types for points and vectors to enforce it.
%(one might say that the standard mathematical formulation of affine
%spaces uses the abstract types approach). 
In this paper we investigate a more sophisticated approach based on 
types indexed by change of origin transformations.

@@

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
we assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The (simplified) index-erasure and relation interpretations
of $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{\rho'} \sepbar \rho' \in
    \extends{\rho}{t\mathord:\SynTransl{2}} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
depends on which (possibly non-compositional) interpretation of
relational environments we use. For now we can think of environments
as mappings from free variables of sort $\SynTransl{2}$ to the group
$\Transl{2}$, and of extension as adding extra components to
mappings. The index-erasure and relational interpretations are given
formally in \autoref{sec:erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\autoref{sec:relational-semantics}. %, respectively.

### 2012-10-24 paper/section-general-framework.tex
various
small changes

Given a particular choice of model and relational interpretation of
primitive types, we can then prove the following.
\begin{theorem}[Abstraction]\label{thm:abstraction}
  If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \semSort{\Delta}$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma} \rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A} \rho$.
\end{theorem}
\begin{proof}
By induction on the typing derivation, making use of
Lemma~\autoref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
Lemma~\autoref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
and Lemma~\autoref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
The details can be found in the Coq development.
\end{proof}

>>>>

Given a particular choice of model and relational interpretation of
primitive types, we can then prove the following.
\begin{theorem}[Abstraction]\label{thm:abstraction}
  If $\Delta; \Gamma \vdash M :
  A$, then for all $\rho \in \semSort{\Delta}$ and $\eta_1, \eta_2
  \in \ctxtSem{\Gamma}$ such that $(\eta_1, \eta_2) \in
  \rsem{\Gamma} \rho$, we have $(\tmSem{M}\eta_1,
  \tmSem{M}\eta_2) \in \rsem{A} \rho$.
\end{theorem}
\begin{proof}
By induction on the typing derivation, making use of
Lemma~\ref{lem:tyeqsubst-relational} (part 1) for rule \TirName{TyEq},
Lemma~\ref{lem:tyeqsubst-relational} (part 2) for rules \TirName{UnivApp} and \TirName{ExPack},
and Lemma~\ref{lem:ctxtsubst-rel} for rules \TirName{UnivAbs} and \TirName{ExUnpack}. 
The details can be found in the Coq development.
\end{proof}

### 2012-10-24 paper/section-introduction.tex
small changes

In previous work the third author studied relational parametricity for
the types parameterized by units-of-measure, whose invariance 
properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
In this paper we investigate the general case of types indexed by attributes
with algebraic structure. For example, in computational geometry, points in
the plane can be indexed by a coordinate frame; in information-flow
security, computations are indexed by principals; in differential privacy, types
can be indexed by `distance'. Types that are
polymorphic over such indices induce invariance properties and
abstraction barriers beyond those introduced by their unindexed
versions, as we now illustrate.

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and abstraction theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

>>>>

In previous work the third author studied relational parametricity for
types parameterized by units-of-measure, whose invariance 
properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
In this paper we investigate the general case of types indexed by attributes
with algebraic structure. For example, in computational geometry, points in
the plane can be indexed by a coordinate frame; in information-flow
security, computations are indexed by principals; in differential privacy, types
can be indexed by `distance'. Types that are
polymorphic over such indices induce invariance properties and
abstraction barriers beyond those introduced by their unindexed
versions, as we now illustrate.

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
